The Unicode Cookbook \linebreak for Linguists

### 2017-12-26 book/unicode.tex
removes some todos
work towards issue 16 - wip

\documentclass[output=inprep,
%  long|short|inprep
%  ,blackandwhite
%  ,smallfont
%  ,draftmode  
		biblatex
		]{LSP/langsci}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
\let\hlipl\hlkwb

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%                                              %%%
%%%             Frontmatter                      %%%
%%%                                              %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\maketitle
\frontmatter
\include{chapters/preface}
% \include{chapters/acknowledgments}
% \include{chapters/abbreviations}
\tableofcontents
\mainmatter%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%                                              %%%
%%%             Chapters                         %%%
%%%                                              %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\include{chapters/introduction}
\include{chapters/pitfalls}
\include{chapters/ipa_background}
\include{chapters/ipa_meets_unicode}
\include{chapters/orthography_profiles}

\chapter{Implementation}
\label{implementation}

To illustrate the practical application of orthography profiles as defined in
the previous Chapter \ref{orthography-profiles}, we have implemented two 
versions of the specifications presented there: one in Python and one
in R.

These two libraries have rather different implementation histories,
and we are not 100\% sure that they will in all situations give the same
results. We do however provide a test suite for each implementation. 
Users should refer to the tests cases and to the documentation in each release 
for specifics about each implementation. Note also that function names differ 
in the two implementations.

Note also that the performance with larger datasets differs, and the code is not
always as clean as we would like it to be. In sum, the two implementations
should be considered as `proof of concept' and not as the final word on the
practical application of the specifications above. In our own experience, the
current implementations are sufficiently fast and stable to be useful for
academic practice (e.g. checking data consistency, or analyzing and
transliterating small to medium sized data sets), but they should probably
better not be used for full-scale industry applications.

This chapter will introduce the implementations, and give practical step-by-step 
guidelines for installing them and using them. Various simple and sometimes 
somewhat abstract examples will be discussed to show the different options 
available, and to illustrate the intended usage of orthography profiles.

This chapter is split into two sections, one for the Python implementation, the 
other for R. Each section has a number of subsections that ...

In addition to the material presented here to get users started, we maintain 
several case studies that use each OP implementation in action. These use cases 
are an the accompanying Github repository available with this book at: 
\url{https://github.com/unicode-cookbook/recipes/}.

% TODO: point to the case studies
% In the next chapter we will then discuss a few full-scale examples to illustrate the application and the usefulness of orthography profiles.

\section{Python}
\label{python-implementations}

The Python package is available both as a command line interface (CLI) 
and as an appplication programming interface (API).

\subsection*{Installation}

To install the Python package from the Python Package Index (PyPI) run:

\begin{lstlisting}[language=bash]
  $ pip install segments
\end{lstlisting}

\noindent on the command line. This will give you access to both 
the CLI and programmatic functionality in Python scripts, when 
you import the \texttt{segments} library.

You can also install the \texttt{segments} 
package with from the Github repository:

\begin{lstlisting}[language=bash]
  $ git clone https://github.com/bambooforest/segments.git
  $ cd segments
  $ python setup.py develop
\end{lstlisting}

\subsection*{Command line interface}

From the command line, access \texttt{segments} and its 
various arguments. For help, run:

\begin{lstlisting}[language=bash]
  $ segments -h
\end{lstlisting}

\noindent Here is an example of how to create and use an orthography 
profile for segmentation. Create a text file:

\begin{lstlisting}[language=bash]
  $ more text.txt
  aäaaöaaüaa
\end{lstlisting}

\noindent Now look at the profile:

\begin{lstlisting}[language=bash,texcl=true]
  $ cat text.txt | segments profile
  Grapheme        frequency       mapping
  a       7       a
  ä       1       ä
  ü       1       ü
  ö       1       ö
\end{lstlisting}

\noindent Write the profile to a file:

\begin{lstlisting}[language=bash]
  $ cat text.txt | segments profile > profile.prf
\end{lstlisting}

\noindent Edit the profile:

\begin{lstlisting}[language=bash]
  $ more profile.prf
  Grapheme        frequency       mapping
  aa      0       x
  a       7       a
  ä       1       ä
  ü       1       ü
  ö       1       ö
\end{lstlisting}

\noindent Now tokenize the text without profile:

\begin{lstlisting}[language=bash]
  $ cat text.txt | segments tokenize
  a ä a a ö a a ü a a	
\end{lstlisting}

\noindent And with profile:
\begin{lstlisting}[language=bash]
  $ cat text.txt | segments --profile=profile.prf tokenize
  a ä aa ö aa ü aa

  $ cat text.txt | segments --mapping=mapping --profile=profile.prf tokenize
  a ä x ö x ü x
\end{lstlisting}

\subsection*{Application programming interface}
The \texttt{segments} API can be accesses by importing 
the package into Python a script. Here is an example of 
how to import the libraries, create a tokenizer object, 
tokenize a string and create a profile.

\begin{lstlisting}[language=python]
>>> from __future__ import unicode_literals, print_function
>>> from segments import Profile, Tokenizer
>>> t = Tokenizer()
>>> t('abcd')
'a b c d'
>>> prf = Profile({'Grapheme': 'ab', 'mapping': 'x'}, \
{'Grapheme': 'cd', 'mapping': 'y'})
>>> print(prf)
mapping Grapheme
ab  x
cd  y
>>> t = Tokenizer(profile=prf)
>>> t('abcd')
'ab cd'
>>> t('abcd', column='mapping')
'x y'
\end{lstlisting}

\section{R}
\label{r-implementation}

% \subsection*{Installing the R implementation}
\subsection*{Installation}

The R implementation is available in the package \texttt{qlcData}, which is 
directly available from the central R repository CRAN (Comprehensive R Archive 
Network). The R software environment itself has to be downloaded from its 
website.\footnote{\url{https://www.r-project.org}} After starting the included 
R program, the \texttt{qlcData} package for dealing with orthography profiles can be 
simply installed as follows:

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# download and install the qlcData software}
\hlkwd{install.packages}\hlstd{(}\hlstr{"qlcData"}\hlstd{)}
\hlcom{# load the software, so it can be used}
\hlkwd{library}\hlstd{(qlcData)}
\end{alltt}
\end{kframe}
\end{knitrout}

The version available through CRAN is the latest stable version.
To obtain the most recent bug-fixes and experimental additions, please use the
development version, which is available on
GitHub.\footnote{\url{http://github.com/cysouw/qlcData}} This development
version can be easily installed using the github-install helper software from the
\texttt{devtools} package.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# download and install helper software}
\hlkwd{install.packages}\hlstd{(}\hlstr{"devtools"}\hlstd{)}
\hlcom{# install the qlcData package from GitHub}
\hlstd{devtools}\hlopt{::}\hlkwd{install_github}\hlstd{(}\hlstr{"cysouw/qlcData"}\hlstd{,} \hlkwc{build_vignettes} \hlstd{=} \hlnum{TRUE}\hlstd{)}
\hlcom{# load the software, so it can be used}
\hlkwd{library}\hlstd{(qlcData)}
\end{alltt}
\end{kframe}
\end{knitrout}

Inside the \texttt{qlcData} package, there are two functions for
orthography processing, \texttt{write.profile} and \texttt{tokenize}. R includes
help files with illustrative examples, and also a so-called `vignette' with
explanations and examples.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# view help files}
\hlkwd{help}\hlstd{(write.profile)}
\hlkwd{help}\hlstd{(tokenize)}
\hlcom{# view vignette with explanation and examples}
\hlkwd{vignette}\hlstd{(}\hlstr{"orthography_processing"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

Basically, the idea is to use \texttt{write.profile} to produce a
basic orthography profile from some data and then \texttt{tokenize} to apply the
(possibly edited) profile on some data, as exemplified in the next section. This
can of course be performed though R, but additionally there are two more
interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
executables and (ii) `shiny' webapps.

The bash executables are little files providing an interface to the R code that
can be used in a shell on a UNIX-alike machine. The location of these
executables is a bit hidden away by the install procedure of R packages. The
location can be found by the following command in R. These executables can be 
used from here, or they can be linked and/or copied to any location as wanted.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# show the path to the bash executables}
\hlkwd{file.path}\hlstd{(}\hlkwd{find.package}\hlstd{(}\hlstr{"qlcData"}\hlstd{),} \hlstr{"exec"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

For example, a good way to use the executables in a terminal is to
make softlinks (using \texttt{ln}) from the executables to a directory in your
PATH, e.g. to \texttt{/usr/local/bin/}. The two executables are named
\texttt{tokenize} and \texttt{writeprofile}, and the links can be made directly 
by using Rscript to get the paths to the executables within the terminal.

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\noindent
\ttfamily
\hlstd{}\hlslc{\#\ get\ the\ paths\ to\ the\ R\ executables\ in\ bash}\hspace*{\fill}\\
\hlstd{pathT}\hlopt{=}\hlstd{}\hlstr{`Rscript\ {-}e\ 'cat(file.path(find.package("qlcData"),\ "exec",\ "tokenize"))'`}\hlstd{}\hspace*{\fill}\\
\hlstd{pathW}\hlopt{=}\hlstd{}\hlstr{`Rscript\ {-}e\ 'cat(file.path(find.package("qlcData"),\ "exec",\ "writeprofile"))'`}\hlstd{}\hspace*{\fill}\\
\hlstd{}\hspace*{\fill}\\
\hlstd{}\hlslc{\#\ make\ softlinks\ to\ the\ R\ executables\ in\ /usr/local/bin}\hspace*{\fill}\\
\hlstd{}\hlkwc{ln\ }\hlstd{}\hlkwb{{-}is\ }\hlstd{}\hlkwd{\$pathT\ \$pathW\ }\hlstd{}\hlopt{/}\hlstd{usr}\hlopt{/}\hlstd{local}\hlopt{/}\hlstd{bin}\hspace*{\fill}
\mbox{}
\normalfont
\end{kframe}
\end{knitrout}

After inserting this softlink it should be possible to access the
\texttt{tokenize} function from the shell. Try \texttt{tokenize --help} to test
the functionality.

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\noindent
\ttfamily
\hlstd{tokenize\ }\hlkwb{{-}{-}help}\hlstd{}\hspace*{\fill}
\mbox{}
\normalfont

\begin{verbatim}
## Error in running command bash
\end{verbatim}
\end{kframe}
\end{knitrout}

To make the functionality even more accessible, we have prepared webapps with 
the \texttt{shiny} framework for the R functions. These webapps are available 
online at \url{TODO}. The webapps are also included inside the \texttt{qlcData} 
package and can be started with the following helper function:

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlkwd{launch_shiny}\hlstd{(}\hlstr{"tokenize"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

% TO DO

\subsection*{Profiles and error reporting}
\label{error-reporting}

The first example of how to use these functions concerns finding errors in the
encoding of texts. In the following example, it looks as if we have two
identical strings, \texttt{AABB}. However, this is just an surface-impression
delivered by the current font, which renders Latin and Cyrillic capitals
identically. We can identify this problem when we produce an orthography profile
from the strings. Using here the R implementation of orthography profiles, we
first assign the two strings to a variable \texttt{test}, and then produce an
orthography profile with the function \texttt{write.profile}. As it turns out,
some of the letters are Cyrillic.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{test} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"AABB"}\hlstd{,} \hlstr{"AАBВ"}\hlstd{)}
\hlkwd{write.profile}\hlstd{(test)}
\end{alltt}
\begin{verbatim}
##   Grapheme Frequency Codepoint                UnicodeName
## 1        A         3    U+0041     LATIN CAPITAL LETTER A
## 2        B         3    U+0042     LATIN CAPITAL LETTER B
## 3        А         1    U+0410  CYRILLIC CAPITAL LETTER A
## 4        В         1    U+0412 CYRILLIC CAPITAL LETTER VE
\end{verbatim}
\end{kframe}
\end{knitrout}

The function of error-message reporting can also nicely be illustrated
with this example. Suppose we made an orthography profile with just the two
Latin letters <A> and <B> as possible graphemes, then this profile would not be
sufficient to tokenize the strings. There are graphemes in the data that are not
in the profile, so the tokenization produces an error, which can be used to fix
the encoding (or the profile). In the example below, we can see that the
Cyrillic encoding is found in the second string of the \texttt{test} input.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{test} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"AABB"}\hlstd{,} \hlstr{"AАBВ"}\hlstd{)}
\hlkwd{tokenize}\hlstd{(test,} \hlkwc{profile} \hlstd{=} \hlkwd{c}\hlstd{(}\hlstr{"A"}\hlstd{,} \hlstr{"B"}\hlstd{))}
\end{alltt}

{\ttfamily\noindent\color{warningcolor}{\#\# Warning in tokenize(test, profile = c("{}A"{}, "{}B"{})): \\\#\# There were unknown characters found in the input data.\\\#\# Check output\$errors for a table with all problematic strings.}}\begin{verbatim}
## $strings
##   originals tokenized
## 1      AABB   A A B B
## 2      AАBВ   A ⁇ B ⁇
## 
## $profile
##   matched_rules Grapheme
## 1             3        B
## 2             3        A
## 
## $errors
##   originals  errors
## 2      AАBВ A ⁇ B ⁇
## 
## $missing
##   Grapheme Frequency Codepoint                UnicodeName
## 1        А         1    U+0410  CYRILLIC CAPITAL LETTER A
## 2        В         1    U+0412 CYRILLIC CAPITAL LETTER VE
\end{verbatim}
\end{kframe}
\end{knitrout}

\subsection*{Different ways to write a profile}
\label{write-profile}

The function \texttt{write.profile} can be used to prepare a skeleton for an
orthography profile from some data. The preparation of an orthography profile
from some data might sound like a trivial problem, but actually there are
various different ways in which strings can be separated into graphemes by
\texttt{write.profile}. Consider the following string of characters called
\texttt{example} below. The default settings of \texttt{write.profile} separates
the string into Unicode graphemes according to grapheme clusters
(`user-perceived characters', see Section~\ref{the-unicode-approach}). As it 
turns out, some of these graphemes are single codepoints, others are combinations
of two codepoints (see Section~\ref{pitfall-characters-are-not-glyphs}).

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{example} \hlkwb{<-} \hlstr{"ÙÚÛÙÚÛ"}
\hlstd{profile_1} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example)}
\end{alltt}
\end{kframe}
\end{knitrout}

% latex table generated in R 3.3.2 by xtable 1.8-2 package
% Thu Feb  2 11:38:04 2017
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ú & 1 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Ù & 1 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
  Û & 1 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 1 (default settings, splitting grapheme clusters)} 
\label{tab:profile1}
\end{table}

By specifying the splitting separator as the empty string
\texttt{sep~=~""}, it is possible to split the string into Unicode codepoints,
thus separating the combining diacritics. The idea behind this option
\texttt{sep} is that separating by a character allows for user-determined
separation. The most extreme choice here is the empty string \texttt{sep~=~""},
which is interpreted as separation everywhere. The other extreme is the default
setting \texttt{sep~=~NULL}, which means that the separation is not
user-defined, but relegated to the Unicode grapheme definitions.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{profile_2} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

% latex table generated in R 3.3.2 by xtable 1.8-2 package
% Thu Feb  2 11:38:04 2017
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
́ & 1 & U+0301 & COMBINING ACUTE ACCENT \\ 
  ̀ & 1 & U+0300 & COMBINING GRAVE ACCENT \\ 
  ̂ & 1 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
  U & 3 & U+0055 & LATIN CAPITAL LETTER U \\ 
  Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 2 (splitting by codepoints)} 
\label{tab:profile2}
\end{table}

Some characters look identical, although they are encoded differently.
Unicode offers different ways of normalization (see
Section~\ref{pitfall-canonical-equivalence}), which can be invoked here as well
using the option \texttt{normalize}. NFC-normalization turns everything into the
precomposed characters, while NFD-normalization separates everything into base
characters with combining diacritics. Splitting by codepoints (i.e. \texttt{sep~=~""}) shows the results of these two normalizations.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# after NFC normalization unicode codepoints have changed}
\hlstd{profile_3} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFC"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
\hlcom{# NFD normalization gives yet another structure of the codepoints}
\hlstd{profile_4} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFD"}\hlstd{,} \hlkwc{sep} \hlstd{=} \hlstr{""}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

% latex table generated in R 3.3.2 by xtable 1.8-2 package
% Thu Feb  2 11:38:04 2017
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 3 (splitting by NFC codepoints)} 
\label{tab:profile3}
\end{table}

% latex table generated in R 3.3.2 by xtable 1.8-2 package
% Thu Feb  2 11:38:04 2017
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
́ & 2 & U+0301 & COMBINING ACUTE ACCENT \\ 
  ̀ & 2 & U+0300 & COMBINING GRAVE ACCENT \\ 
  ̂ & 2 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
  U & 6 & U+0055 & LATIN CAPITAL LETTER U \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 4 (splitting by NFD codepoints)} 
\label{tab:profile4}
\end{table}

It is important to realize that for Unicode grapheme definitions, NFC
and NFD normalization are equivalent. This can be shown by normalizing the
example in either NFC or NFD, but using default separation in
\texttt{write.profile}. To be precise, default separation means setting
\texttt{sep~=~NULL}, but that has not be added explicitly.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# note that NFC and NFD normalization are identical}
\hlcom{# for unicode grapheme definitions}
\hlstd{profile_5} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFD"}\hlstd{)}
\hlstd{profile_6} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFC"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

% latex table generated in R 3.3.2 by xtable 1.8-2 package
% Thu Feb  2 11:38:04 2017
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
  Ù & 2 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
  Û & 2 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 5 (splitting by graphemes after NFD)} 
\label{tab:profile5}
\end{table}

% latex table generated in R 3.3.2 by xtable 1.8-2 package
% Thu Feb  2 11:38:04 2017
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 6 (splitting by graphemes after NFC)} 
\label{tab:profile6}
\end{table}

Note that these different profiles can also be produced using the bash
executable \texttt{writeprofile} (see \ref{r-implemenation} for notes
on how to install the bash executable). Exactly this example is also included in
the help file of the executable.

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\noindent
\ttfamily
\hlstd{writeprofile\ }\hlkwb{{-}{-}help}\hlstd{}\hspace*{\fill}
\mbox{}
\normalfont

\begin{verbatim}
## Error in running command bash
\end{verbatim}
\end{kframe}
\end{knitrout}

\subsection*{Using an orthography profile skeleton}
\label{profile-skeleton}

A common workflow to use these functions is to first make a skeleton for an
orthography profile and then edit this profile by hand. For example, Table
\ref{tab:profile_skeleton1} shows the profile skeleton after a few graphemes have
been added to the file. Note that in this example, the profile is written to the
desktop, and this file has to be edited manually. We simply add a few
multigraphs to the column \texttt{Grapheme} and leave the other columns empty.
these new graphemes are then included in the graphemic parsing.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# a few words to be graphemically parsed}
\hlstd{example} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"mishmash"}\hlstd{,} \hlstr{"mishap"}\hlstd{,} \hlstr{"mischief"}\hlstd{,} \hlstr{"scheme"}\hlstd{)}
\hlcom{# write a profile skeleton to a file}
\hlkwd{write.profile}\hlstd{(example,} \hlkwc{file} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}\hlstd{)}
\hlcom{# edit the profile, and then use the edited profile to tokenize}
\hlkwd{tokenize}\hlstd{(example,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}\hlstd{)}\hlopt{$}\hlstd{strings}
\end{alltt}
\end{kframe}
\end{knitrout}

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
##   originals    tokenized
## 1   shampoo  sh a m p oo
## 2    mishap   m i sh a p
## 3  mischief m i sch ie f
## 4    scheme    sch e m e
\end{verbatim}
\end{kframe}
\end{knitrout}

% latex table generated in R 3.3.2 by xtable 1.8-2 package
% Thu Feb  2 11:38:04 2017
\begin{table}[htb]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & UnicodeName \\ 
  \midrule
sh &  &  &  \\ 
  ch &  &  &  \\ 
  sch &  &  &  \\ 
  ie &  &  &  \\ 
  oo &  &  &  \\ 
  a & 2 & U+0061 & LATIN SMALL LETTER A \\ 
  c & 2 & U+0063 & LATIN SMALL LETTER C \\ 
  e & 3 & U+0065 & LATIN SMALL LETTER E \\ 
  f & 1 & U+0066 & LATIN SMALL LETTER F \\ 
  h & 4 & U+0068 & LATIN SMALL LETTER H \\ 
  i & 3 & U+0069 & LATIN SMALL LETTER I \\ 
  m & 4 & U+006D & LATIN SMALL LETTER M \\ 
  o & 2 & U+006F & LATIN SMALL LETTER O \\ 
  p & 2 & U+0070 & LATIN SMALL LETTER P \\ 
  s & 4 & U+0073 & LATIN SMALL LETTER S \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Manually edited profile skeleton} 
\label{tab:profile_skeleton1}
\end{table}

It is also possible to leave out the Unicode information in
the profile skeleton by using the option \texttt{info = FALSE}. It is also
possible not to use a separate file at all, but process everything within R. In
simple situations this is often useful (see below), but in general we prefer to
handle everything through a separately saved orthography profile. This profile
often contains highly useful information that is nicely coded and saved inside
this one file, and can thus be easily distributed and shared. Doing the same as
above completely within R might look as follows:

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# make a profile, just select the column 'Grapheme'}
\hlstd{profile} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example)[,} \hlstr{"Grapheme"}\hlstd{]}
\hlcom{# extend the profile with multigraphs}
\hlstd{profile} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"sh"}\hlstd{,} \hlstr{"ch"}\hlstd{,} \hlstr{"sch"}\hlstd{,} \hlstr{"ie"}\hlstd{,} \hlstr{"oo"}\hlstd{, profile)}
\hlcom{# use the profile to tokenize}
\hlkwd{tokenize}\hlstd{(example, profile)}\hlopt{$}\hlstd{strings}
\end{alltt}
\begin{verbatim}
##   originals    tokenized
## 1   shampoo  sh a m p oo
## 2    mishap   m i sh a p
## 3  mischief m i sch ie f
## 4    scheme    sch e m e
\end{verbatim}
\end{kframe}
\end{knitrout}

\subsection*{Rule ordering}
\label{rule-ordering}

Not yet everything is correct in this graphemic parsing of the example discussed
previously. The sequence <sh> in `mishap' should not be a digraph, and
conversely the sequence <sch> in `mischief' should of course be separated into
<s> and <ch>. One of the important issues to get the graphemic parsing right is
the order in which graphemes are parsed. For example, currently the grapheme
<sch> is parsed before the grapheme <ch>, leading to <m\ i\ sch\ ie\ f> instead
of the intended <m\ i\ s\ ch\ ie\ f>. The reason that <sch> is parsed before
<ch> is that by default longer graphemes are parsed before shorter ones. Our
experience is that in most cases this is expected behaviour. You can change the
ordering by specifying the option \texttt{ordering}. Setting this option to
\texttt{NULL} results in no preferential ordering, i.e. the graphemes are parsed
in the order of the profile, from top to bottom. Now `mischief' is parsed
correctly, but `scheme' is wrong. So this ordering is not the solution in this
case.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# do not reorder the profile}
\hlcom{# just apply the graphemes from top to bottom}
\hlkwd{tokenize}\hlstd{( example}
         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
         \hlstd{,} \hlkwc{ordering} \hlstd{=} \hlkwa{NULL}
        \hlstd{)}\hlopt{$}\hlstd{strings}
\end{alltt}
\end{kframe}
\end{knitrout}

>>>>

\documentclass[output=inprep,
%  long|short|inprep
%  ,blackandwhite
%  ,smallfont
%  ,draftmode  
		biblatex
		]{LSP/langsci}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%                                              %%%
%%%             Frontmatter                      %%%
%%%                                              %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\maketitle
\frontmatter
\include{chapters/preface}
% \include{chapters/acknowledgments}
% \include{chapters/abbreviations}
\tableofcontents
\mainmatter%

@@
